---
title: "Darwin Core mapping"
subtitle: "For: List of Invasive Alien Species of Union concern"
author:
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

Install required libraries (only if the libraries have not been installed before):

```{r}
installed <- rownames(installed.packages())
required <- c(
  "tidyverse",
  "tidylog",
  "magrittr",
  "here",
  "janitor",
  "digest"
)
if (!all(required %in% installed)) {
  install.packages(required[!required %in% installed])
}
```

Load libraries:

```{r message = FALSE}
library(tidyverse)      # To do data science
library(tidylog)        # To provide feedback on dplyr functions
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(digest)         # To generate hashes
```

# Read source data

Create a data frame `input_data` from the source data:

```{r}
input_data <- read_tsv(
  file = here("data", "raw", "list_of_invasive_alien_species_of_union_concern.tsv"),
  show_col_types = FALSE
)
```

Preview data:

```{r}
input_data %>% head(n = 5)
```

# Process source data

## Tidy data

The column `scientific_name_alternative` contains recent taxonomic changes. We remove this column by making new rows for it and copying the information related to English vernacular name (column `english_name`), entry into force date (column `entry_into_force`) and kingdom (column `kingdom`):

```{r}
input_data %<>% 
  pivot_longer(
    starts_with("scientific_name"),
    names_to = "col_name", # help column
    values_to = "scientific_name"
  ) %>%
  select(-col_name) %>%
  relocate(scientific_name) %>%
  filter(!is.na(scientific_name))
```

This operation results in the following data.frame:

```{r}
input_data
```

## Taxon IDs

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`. Here we create one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name and kingdom (that will remain the same as long as scientific name and kingdom remain the same):

```{r}
vdigest <- Vectorize(digest) # Vectorize digest function to work with vectors
input_data %<>% mutate(taxon_id = paste(
  "union-list",
  "taxon",
  vdigest(paste(scientific_name, kingdom), algo = "md5"),
  sep = ":"
))
```

## Preview data

Show the number of taxa and distributions per kingdom:

```{r}
input_data %>%
  group_by(kingdom) %>%
  summarize(
    `# taxa` = n_distinct(taxon_id),
    `# rows` = n()
  ) %>%
  adorn_totals("row")
```

Preview data:

```{r}
input_data %>% head()
```

# Taxon core

Create a `taxon` dataframe starting from `input_data`:

```{r}
taxon <- input_data
```

## Term mapping

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

Start with record-level terms which contain metadata about the dataset (which is generally the same for all records).

### language

```{r}
taxon %<>% mutate(dwc_language = "en")
```

### license

```{r}
taxon %<>% mutate(dwc_license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(dwc_rightsHolder = "INBO")
```

### datasetID

```{r}
taxon %<>% mutate(dwc_datasetID = "")
```

### institutionCode

```{r}
taxon %<>% mutate(dwc_institutionCode = "INBO")
```

### datasetName

```{r}
taxon %<>% mutate(dwc_datasetName = "List of Invasive Alien Species of Union concern")
```

The following terms contain information about the taxon:

### taxonID

```{r}
taxon %<>% mutate(dwc_taxonID = taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(dwc_scientificName = scientific_name)
```

### kingdom

Inspect values:

```{r}
taxon %>%
  group_by(kingdom) %>%
  count()
```

Map values:

```{r}
taxon %<>% mutate(dwc_kingdom = kingdom)
```

### taxonRank

Map values by recoding to the [GBIF rank vocabulary](http://rs.gbif.org/vocabulary/gbif/rank_2015-04-24.xml):

```{r}
taxon %<>% mutate(dwc_taxonRank = case_when(
  # Scientific names often end with authors, so regex only considers beginning
  scientific_name == "Lespedeza juncea var. sericea (Thunb.) Lace & Hauech" ~ "variety",
  scientific_name == "Lithobates (Rana) catesbeianus Shaw, 1802" ~ "species",
  scientific_name == "Procambarus fallax (Hagen, 1870) f. virginalis" ~ "form",
  scientific_name == "Pueraria montana (Lour.) Merr. var. lobata (Willd.)" ~ "variety",
  scientific_name == "Vespa velutina nigrithorax de Buysson, 1905" ~ "subspecies",
  str_detect(scientific_name, "^[A-Z][a-z]+ [a-z]+") ~ "species"
))
```

Show mapped values: 

```{r}
taxon %>%
  filter(is.na(dwc_taxonRank)) %>%
  select(dwc_scientificName)
```

## Post-processing

Only keep the Darwin Core columns:

```{r}
taxon %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(taxon) <- str_remove(colnames(taxon), "dwc_")
```

Preview data:

```{r}
taxon %>% head()
```

Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
```

# Map vernacular names extension

## Pre-processing

Create a `vernacular` dataframe starting from `input_data`:

```{r}
vernacular <- input_data
```

## Term mapping

Map the data to [Vernacular Names](http://rs.gbif.org/extension/gbif/1.0/vernacularname.xml).

### TaxonID

```{r}
vernacular %<>% mutate(dwc_taxonID = taxon_id)
```

### vernacularName

```{r}
vernacular %<>% mutate(dwc_vernacularName = english_name)
```

### language

Set language code to `en` (English):

```{r}
vernacular %<>% mutate(dwc_language = "en")
```

## Post-processing

Only keep the Darwin Core columns:

```{r}
vernacular %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(vernacular) <- str_remove(colnames(vernacular), "dwc_")
```

Preview data:

```{r}
vernacular %>% head()
```

Save to CSV:

```{r}
write_csv(vernacular, here("data", "processed", "vernacularname.csv"), na = "")
```
